Synthesis of production rules for glycans

### 2020-10-15 papers/sugar-synth/algo.tex

\begin{wrapfigure}{r}{0.37\textwidth}
    \vspace{-6mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position may be in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position may not be present in the rule.
The dashed area includes absent nodes
from the full tree.
Among the matching nodes, we make two cases, namely if the expanding
nodes are the descendants of the matching node or not. 
The third situation is when a node position is in the matching part
and has expanding decadents.
The solid path from $v_0$ to the root $v_e$ of the expanding part includes
the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is when a node position is in the matching part and
has no expanding decedents.
The nodes in the rest of the light gray area are in the fourth situation.
Another variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
We will use the pool of $KVars$ to add variables to the templates.
Formally, the templates are defined as follows.

\begin{df}
For given integers $d$ and $w$,
a {\em rule template} $t = (V, \nu, \kappa, C, v_{0r})$ is a labeled full tree with depth $d$ and 
each internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to distinct
sugar choice variables, 
$\kappa : V \maps KVars$ maps nodes to distinct situation variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0r} \in V$ is the root of the tree.
\end{df}
In the method at line~\ref{line:createRtemp}, we use the
call to \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
height $d$ and children width $w$. We get $d$ and $n$ as input, and
we let $w$ be the maximum arity of any sugar that occurs in $\mu$.
This allows us to map any node to any sugar.
We also assume that templates
returned by $\textsc{MakeTemplatesRule}$ do not share variables.
We obtain a production rule from a template by assigning values
to the variables.
If for a node $v$ in a template if we assign $\kappa(v) = Absent$,
we call the node {\em absent}. Otherwise, we call the node {\em present}.
Not all the assignments may result in a valid rule.
Next at line~\ref{line:ruleCorr}, we will construct constraints
that encode the set of valid rules.

A valid assignment to the variables in
a template  $t = (V, \nu, \kappa, C, r)$ must
satisfy the following six conditions.
The first two encode the basic agreement among variables.
The third encodes the tree structure of the rule.
The last three encode the relative positions of the four
parts of the rule.
\begin{enumerate}
\item If a node is present, then it is labeled with a sugar.
  %Otherwise, we ignore the label.
$$
\Land_{s \in S} \Land_{v \in V} (\kappa(v) \neq Absent \limplies \nu(v) \neq \bot )
$$
\item  The number of children of a node should match with the arity of the labeled sugar.
  The following formula states that the children that are at greater arity than that of the label
  must be absent.
$$
\Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
$$
\item If a node is present, then the parent of the node is also present.
  $$
  \Land_{{\text{internal node } v \in V}}
  \Land_{i\in[1,w]} (\kappa(C(v,i) ) \neq Absent \limplies \kappa( v ) \neq Absent )  
  $$
\item If a node is $Expand$, then its children are also $Expand$ if present. The following constraints state that once one reached the expanding part, there
are no descendants that are matching.
  $$
\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Expand \limplies \kappa(C(v,i)) \in \{Expand,Absent\}   )
$$
\item If a node is $Match$, then its children are also $Match$ if present.
  Similarly, the following constraints state that once one reached the matching part without expanding descendants, there are no descendants that are expanding.
$$
\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Match \limplies \kappa(C(v,i)) \in \{Match,Absent\}   )
$$
\item If a node is $MatchAns$, then exactly one child is $MatchAns$ or $Expand$.
  The following constraints encode that there is exactly one subtree that is expanding part
  by stating that there is no branching in $MatchAns$ nodes and they form a path.  
$$
\Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = MatchAns \limplies \sum_{i\in[1,w]}( \kappa(C(v,i)) \in \{MatchAns,Expand\}) = 1
$$
\end{enumerate}

In the method, the call to $\textsc{RuleTemplateCorrectness}$ at
line~\ref{line:ruleCorr} creates the above constraints
and store them in $tCons$.
Next, we will also be searching for the molecules that may be produced by the learned rules.
Therefore, we need to define the search space for the molecules.
We use similar templates for defining the search space.
Since we always search in bounded space, we limit the template size using a parameter,
namely the height of the template.
\begin{df}
For given integers $h$ and $w$,
a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
every internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0m} \in V$ is the root of the tree.
\end{df}
At line~\ref{line:createMtemp}, we use the call to \textsc{MakeTemplateMol}($S$, $h$, $w$) to
create a template of height $h$ and children width $w$.
%
Our method searches for unwanted producible molecules up to the height of $h$,
which we choose to be the maximum of the heights of molecules in $\mu$.
%
The choice of $h$ is somewhat arbitrary.
%
We assume if the rules could have produced a molecule with height less than $h$,
then it must be in $\mu$.
On the other hand, we ignore the potential formation of the large size of molecules.
%
Since we observe only finite molecules in any wet experiment, we may assume
there may be large producible molecules, but we did not observe them.
However, this point is debatable and $h$ may be viewed as a parameter of the whole
method.

%
Similar to the rule templates, not all assignments to molecule template variables
are valid.
We add the following four conditions for
valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
The first  two conditions encode the tree structure of the molecule and
matching the number of children with label arity for each node.
The second encodes that the height of the molecule is large enough.
The last encodes that new molecule must not be the same as any molecule in $\mu$.
\begin{enumerate}
\item If a node is present, then the parent of the nodes is also present.
  $$
  \Land_{i\in[1,w]}\Land_{\mathrlap{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
  $$
\item The children count of a node matches with the arity of the labeled sugar.
  $$
  \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
  $$
\item At least one of the leaves is present. This is for
  avoiding the trivial solution.
  $$
  \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
  $$
  where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
  This allows us to look for all molecules that are close
  to height $h$.
  We choose $d' = h-d$, where $d$ is the height of the rule templates.
  This $d'$ ensures that we do not miss producible molecules that
  can not have exact height $h$ but is slightly larger than $h$.
  Then, there must be another producible
  molecule that is a subtree of the molecule and has height in
  between $h$ and $h-d$
  because the maximum depth of a rule is $d$.
\item We find a molecule that is not in $\mu$.
  To encode the constraints, let predicate $Neq$ be recursively defined as follows.
  \begin{align*}
  Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
  \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
    Neq(v,\bot) :=\;&  \nu(v) \neq \bot
  \end{align*}
  The constraint is
  % \begin{align*}
  $\Lor_{%\mathclap
    {m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$
  % \end{align*}
  In our implementation, 
  we have an option to also ignore the subtrees of
  the molecules of $\mu$.
\end{enumerate}
%
In our method, the call to $\textsc{MolTemplateCorrectness}$ at
line~\ref{line:molCorr} implements above constraints and store them in
$mCons$.

We need to encode that the rules do generate the molecules
in $\mu$ and do not generate any other.
Using procedure $\textsc{EncodeProduce}$, we generate
the corresponding constraints.
We will discuss the procedure in-depth shortly.
Let us continue with~\textsc{SugarSynth}.
At~\ref{line:molenc} line, we call $\textsc{EncodeProduce}$ for
each molecule in $\mu$ and generate constraints $pCons$ stating that
the solutions of the templates will produce the molecules in $\mu$.

After producing the constraints $tCons$, $mCons$, and $pCons$,
the method enters in the loop.
It first checks the satisfiability of conjunction
$tCons \land pCons \land nCons$
using some SMT solver, where $nCons$ is $\ltrue$ in the first iteration
and will encode constraints related to counterexample molecules.
If the conjunction is unsatisfiable, there are no rules of
the input number and height,
and the method returns failure of synthesis.
If it is satisfiable, it constructs the rules at line~\ref{line:getR} from
the satisfying assignment $a$
and stores them in $R$.

At line~\ref{line:consNewR}, we construct constraints $rCons$ again using
$\textsc{EnocdeProduce}$ that says that template molecule $\hat{m}$
is generated by rules $R$.
We check the satisfiability of $rCons \land mCons$.
If it is not satisfiable, we have found the rules that generate exactly
the molecules in $\mu$ within our search range of counterexample molecules,
and the loop terminates.
Otherwise, we use the satisfying assignment to create a
counterexample molecule $m'$.
At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
all possible applications of templates rules in $T$ must not produce $m'$.
We use shorthand $F \landplus G$ for $ F := F \land G$.
As we will see that $\textsc{EnocdeProduce}$ introduces fresh variable
maps $\tau$ and $cuts$ in the encoding.
Since we negate the returned formula by $\textsc{EnocdeProduce}$ 
and then we check the satisfiability, the encoding will not be correct
due to the fresh variables.
Therefore, we will need to introduce universal quantifiers to the
fresh variables.
We introduce universal quantifiers over $\tau$ and $cuts$
variables before adding to $nCons$.
Intuitively, the universal quantifiers say that all possible
application patterns of the learned rules do not produce $m'$.
Afterword, the loop goes to the next iteration.
Since the domain of all the variables is finite, eventually the loop must
terminate.

@@

Now let us look at the encoding generated by $\textsc{EncodeProduce}$
in detail.
The process of production of molecules adds pieces of trees of nodes
one after another.
In order to show that a molecule is producible by a set of production rules,
we need to find the nodes where the production rules are applied,
the rules that are applied on the nodes, and
the order of the application of the rules.
The encoding is like encoding bounded executions of programs.

To model the production due to the application of the rules,
we need to attach three maps to the nodes of the molecules.
\begin{itemize}
\item Let $cuts$ map each node to a Boolean variable
indicating the node is the point where a rule is applied and the
molecule was expanded.
We say points of the applications of the production
rules as {\em cuts} of the tree.
All the nodes that are in between the cuts are added in a single step.

\item Let $rmatch$ map each node to a rule indicating
  the rule that is applied to expand at the node.
  We need to match a rule to a node if it is a cut point.
  % If a node is cut point, we need to the rules that added the nodes  
\item Let $\tau$ map each node to an integer variable
  indicating the time point when the node was added to the molecule.
  The order of the addition is an important constraint since already
  existing nodes in a molecule decide what can be added in the future.
  If we do not capture the order in our encoding, we may wrongly declare
  that a molecule is producible while it is not due to the matching
  constraints.
\end{itemize}

Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
that returns the encoding.
It takes a molecule $m$ and a set of rules $T$.
Both the inputs can be template or concrete.
The function is tolerant of the variations, but our presentation
assumes that the molecule is concrete and the rules are templates.
This will cover the case when the $\textsc{EncodeProduce}$ is called at
line~\ref{line:molenc} in algorithm~\ref{alg:sugar-synth}.
However, at line \ref{line:consNewR} the molecule is a template
and the rules are concrete.
We will discuss the handling of this situation at the end of this subsection.
$\textsc{EncodeProduce}$ is implemented with the help of three other functions,
$\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
All of them are presented in algorithm~\ref{alg:produce}.
%

$\textsc{EncodeProduce}$ returns constraints stating for each node $v$
and rule template $t$ that
if $v$ is at a cut and $t$ is applied at $v$, then $t$ must match
at the node $v$.
We require $rmatch(v)$ to be equal to some rule.
Therefore, if $v$ is a cut node, some template must be applied. 
Since we are matching with template rule, we do not know the
position of the head of expanding nodes.
We enumerate to all possible depths from $1$ to $d-1$ for finding
the head.
For each $\ell \in [1,d)$, we call $\textsc{EncodeP}( v, t, \ell)$
to construct the constraints encoding that $t$ is applied at $v$
and the depth of the root of the expanding nodes is at depth $\ell$.

In $\textsc{EncodeP}$,
we need to match rule template $t$ such that the root of the expanding
nodes matches with $v$.
Since we are given a guess $\ell$ for the depth of the
node.
We can traverse up from $v$ for $\ell$ steps to find the node to match
the root of $t$.
Naturally, the $\textsc{EncodeP}$ ignores returns $\lfalse$ if there
is no $\ell$th ancestor of $v$, i.e., there is no possibility of a match.
The variable $mark$ is assigned the timestamp for node $v$.
During matching, we will add constraints stating that the matching part of the
template should match to the nodes added before $mark$
and expanding part of the template should match with nodes that
are added at or after $mark$.
The local variable $c$ collects a conjunction of constraints as they are
generated in the function and $c$ is initially $\ltrue$.
The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
and traverses the nodes in $t$.
The while loop at~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
the $\ell$th ancestor from $v$,
matches all the ancestors up to the parent of $v$.
As loop traverses down, it also traverses $t$ along the matching
using variable $v_r$.
In each iteration, $v_r$ is updated to the $j$th child due to lines
\ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
if the ancestors of $v$ also traverse along with the $j$th child
(see if condition at line \ref{line:encodep-jth-child-cond}).

Let us consider the body of the loop in detail.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
declared $MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The for loop at \ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
$\textsc{MatchTree}$ is also called with a third parameter
$mark$ and a Boolean value as the fourth parameter.
We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get the node name $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
After the while loop in $\textsc{EncodeP}$, $v_r$ is the
candidate node for the root of expanding nodes.
At line \ref{line:encodep-vr-expand}, we declare that $v_r$
is an $Expand$ node.
Afterword, we match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$ again at line~\ref{line:encodep-mtree-expand}.
At line~\ref{line:mcut}, we also call to match the cut pattern at
the subtree of $v$ with the template rule.

>>>>

\begin{wrapfigure}{r}{0.37\textwidth}
    \vspace{-12mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes.
Among the matching nodes, we have two cases.
% , namely if the expanding
% nodes are the descendants of the matching node or not.
The third situation is when a node position is in the matching part
and has expanding decadents.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
The templates are defined as follows.

\begin{df}
For given integers $d$ and $w$,
a {\em rule template} $t = (V, \nu, \kappa, C, v_{0r})$ is a labeled full tree with depth $d$ and 
each internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to distinct
sugar choice variables, 
$\kappa : V \maps KVars$ maps nodes to distinct situation variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0r} \in V$ is the root of the tree.
\end{df}
In the algorithm at line~\ref{line:createRtemp}, we call \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
height $d$ and children width $w$.
% We get $d$ and $n$ as input, and
% we let $w$ be the maximum arity of any sugar that occurs in $\mu$.
Since $w$ is the maximum arity of any sugar, we can map any node to any sugar.
% We also assume that templates
% returned by $\textsc{MakeTemplatesRule}$ do not share variables.
% We obtain a production rule from a template by assigning values
% to the variables of the template.
For a node $v$ in a template if we assign $\kappa(v) = Absent$,
we call the node {\em absent}. Otherwise, we call the node {\em present}.
% Not all the assignments may result in a valid rule.
Next at line~\ref{line:ruleCorr}, we will construct constraints
that encode the set of valid rules. 
A valid assignment to the variables in
a template  $t = (V, \nu, \kappa, C, r)$ must
satisfy the following six conditions.
% The first two encode the basic agreement among variables.
% The third encodes the tree structure of the rule.
% The last three encode the relative positions of the four
% parts of the rule.
\begin{enumerate}
\item If a node is present, then it is labeled with a sugar.\\
  %Otherwise, we ignore the label.
$
\Land_{s \in S} \Land_{v \in V} (\kappa(v) \neq Absent \limplies \nu(v) \neq \bot )
$
\item
  % The number of children of a node should match with the arity of the labeled sugar.
  % The following formula states that

  The children that are at greater arity than that of the label
  are absent.\\
$
\Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
$
\item If a node is present, then the parent of the node is also present.\\
  $
  \Land_{{\text{internal node } v \in V}}
  \Land_{i\in[1,w]} (\kappa(C(v,i) ) \neq Absent \limplies \kappa( v ) \neq Absent )  
  $
\item If a node is $Expand$, then its children are also $Expand$ if present.\\
  % The following constraints state that once one reached the expanding part, there are no descendants that are matching.
  $
\Land_{i\in[1,w]} \Land_{{\text{internal node } v \in V}} (\kappa(v) = Expand \limplies \kappa(C(v,i)) \in \{Expand,Absent\}   )
$
\item If a node is $Match$, then its children are also $Match$ if present.\\
  % Similarly, the following constraints state that once one reached the matching part without expanding descendants, there are no descendants that are expanding.
$
\Land_{i\in[1,w]} \Land_{{\text{internal node } v \in V}} (\kappa(v) = Match \limplies \kappa(C(v,i)) \in \{Match,Absent\}   )
$
\item If a node is $MatchAns$, then exactly one child is $MatchAns$ or $Expand$.\\
  % The following constraints encode that there is exactly one subtree that is expanding part
  % by stating that there is no branching in $MatchAns$ nodes and they form a path.
\mbox{}\hspace{-5mm}  
$
\Land_{v \in V} (\kappa(v) = MatchAns \limplies \sum_{i\in[1,w]}( \kappa(C(v,i)) \in \{MatchAns,Expand\}) = 1
$
\end{enumerate}

The call to $\textsc{RuleTemplateCorrectness}$ at
line~\ref{line:ruleCorr} creates the above constraints
and store them in $tCons$.
Next, we will also be searching for the molecules that may be produced by the learned rules.
Therefore, we need to define the search space for the molecules.
We use templates for defining the search space.
We limit the template size using a parameter,
namely the height of the template.
\begin{df}
For given integers $h$ and $w$,
a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
each internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0m} \in V$ is the root.
\end{df}
At line~\ref{line:createMtemp}, we use the call to \textsc{MakeTemplateMol}($S$, $h$, $w$) to
create a template of height $h$ and children width $w$.
%
Our method searches for unwanted producible molecules up to the height of $h$,
which we set to the maximum of the heights of molecules in $\mu$.
%
The choice of $h$ is arbitrary.
% %
% We assume if the rules could have produced a molecule with height less than $h$,
% then it must be in $\mu$.
% On the other hand, 
% We ignore the potential formation of the large size of molecules.
% %
% Since we observe only finite molecules in any wet experiment, we may assume
% there may be large producible molecules, but we did not observe them.
% However, this point is debatable and $h$ may be viewed as
% a parameter of the whole method.

%
Similar to the rule templates, not all assignments to molecule template variables
are valid.
We add the following conditions for
valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
% The first  two conditions encode the tree structure of the molecule and
% matching the number of children with label arity for each node.
% The second encodes that the height of the molecule is large enough.
% The last encodes that new molecule must not be the same as any molecule in $\mu$.
\begin{enumerate}
\item If a node is present, then the parent of the nodes is also present.\\
  $
  \Land_{i\in[1,w]}\Land_{{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
  $
\item The children count of a node matches with the arity of the labeled sugar.\\
  $
  \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
  $
% \item At least one of the leaves is present. This is for
%   avoiding the trivial solution.\\
%   $
%   \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
%   $
%   where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
%   This allows us to look for all molecules that are close
%   to height $h$.
%   We choose $d' = h-d$, where $d$ is the height of the rule templates.
  % This $d'$ ensures that we do not miss producible molecules that
  % can not have exact height $h$ but is slightly larger than $h$.
  % Then, there must be another producible
  % molecule that is a subtree of the molecule and has height in
  % between $h$ and $h-d$
  % because the maximum depth of a rule is $d$.
\item We find a molecule that is not in $\mu$.
  We encode $\Lor_{{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$, where
   predicate $Neq$ be recursively defined as follows.\\
  $Neq(v, v') :=\; \nu(v) \neq M(v') \lor
  \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) )$\\
  $Neq(v,\bot) :=\;  \nu(v) \neq \bot$
  % \begin{align*}
  % Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
  % \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
  %   Neq(v,\bot) :=\;&  \nu(v) \neq \bot
  % \end{align*}
  % The constraint is
  % % \begin{align*}
  % % \end{align*}
  % In our implementation, 
  % we have an option to also ignore the subtrees of
  % the molecules of $\mu$.
\end{enumerate}
%
In our method, the call to $\textsc{MolTemplateCorrectness}$ at
line~\ref{line:molCorr} implements above constraints and store them in
$mCons$.

We need to encode that the rules do generate the molecules
in $\mu$ and do not generate any other.
Using procedure $\textsc{EncodeProduce}$, we generate
the corresponding constraints.
We will discuss the procedure in-depth shortly.
Let us continue with~\textsc{SugarSynth}.
At line~\ref{line:molenc}, we call $\textsc{EncodeProduce}$ for
each molecule in $\mu$ and generate constraints $pCons$ stating that
the solutions of the templates will produce the molecules in $\mu$.

After producing the constraints $tCons$, $mCons$, and $pCons$,
the method enters in the loop.
It first checks the satisfiability of conjunction
$tCons \land pCons \land nCons$, where $nCons$ is $\ltrue$ in the first iteration
and will encode constraints related to counterexample molecules.
If the conjunction is unsatisfiable, there are no rules of
the input number and height,
and the method returns failure of synthesis.
If it is satisfiable, it constructs the rules at line~\ref{line:getR} from
%the satisfying assignment $a$
and stores them in $R$.

At line~\ref{line:consNewR}, we construct constraints $rCons$ again using
$\textsc{EnocdeProduce}$ that says that template molecule $\hat{m}$
is generated by rules $R$.
We check the satisfiability of $rCons \land mCons$.
If it is not satisfiable, we have found the rules that generate exactly
the molecules in $\mu$ and the loop terminates.
Otherwise, we use the satisfying assignment to create a
counterexample molecule $m'$.
At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
all possible applications of template rules in $T$ must not produce $m'$.
We use shorthand $F \landplus G$ for $ F := F \land G$.
As we will see that $\textsc{EnocdeProduce}$ introduces fresh variable
maps $\tau$ and $cuts$ in the encoding.
Since we negate the returned formula by $\textsc{EnocdeProduce}$ 
and then we check the satisfiability.
% the encoding will not be correct
% due to the fresh variables.
Therefore, we need to introduce universal quantifiers to the
fresh variables.
We introduce universal quantifiers over $\tau$ and $cuts$
variables before adding to $nCons$.
% Intuitively, the universal quantifiers say that all possible
% application patterns of the learned rules do not produce $m'$.
Next, the loop goes to the next iteration.
Since the domain of all the variables is finite, eventually the loop must
terminate.

@@

Now let us look at the encoding generated by $\textsc{EncodeProduce}$.
The process of production of molecules adds pieces of trees of nodes
one after another.
In order to show that a molecule is producible by a set of production rules,
we need to find the nodes where the production rules are applied,
the rules that are applied on the nodes, and
the order of the application of the rules.
% The encoding is like encoding bounded executions of programs.
To model the production due to the application of the rules,
we attach three maps to the molecule nodes.
\begin{itemize}
\item Let $cuts$ map each node to a Boolean variable
indicating the node is the point where a rule is applied to
expand the molecule.
We say points of the applications of the production
rules as {\em cuts} of the tree.
% All the nodes that are in between the cuts are added in a single step.

\item Let $rmatch$ map each node to a rule indicating
  the rule that is applied to expand at the node.
  We need to match a rule to a node if it is a cut point.
  % If a node is cut point, we need to the rules that added the nodes  
\item Let $\tau$ map each node to an integer variable
  indicating the time point when the node was added to the molecule.
  Since already added nodes in a molecule decide what can be added later,
  we need to record the order of the addition.
  % If we do not capture the order in our encoding, we may wrongly declare
  % that a molecule is producible while it is not due to the matching
  % constraints.
\end{itemize}

Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
that returns the encoding.
It takes a molecule $m$ and a set of rules $T$.
Both the inputs can be template or concrete.
% The function is tolerant of the variations, but
Our presentation assumes that the molecule is concrete and
the rules are templates.
This will cover the case when the $\textsc{EncodeProduce}$ is called at
line~\ref{line:molenc} in algorithm~\ref{alg:sugar-synth}.
However, at line \ref{line:consNewR} the molecule is a template
and the rules are concrete, which we will discuss later.
% We will discuss the handling of this situation at the end of this subsection.
$\textsc{EncodeProduce}$ uses help of three other supporting functions,
$\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
% All of them are presented in algorithm~\ref{alg:produce}.
%

$\textsc{EncodeProduce}$ returns constraints stating for each node $v$
and rule template $t$,
if $v$ is at a cut and $t$ is applied at $v$, then $t$ must match
at the node $v$.
We require $rmatch(v)$ to be equal to some rule.
% Therefore, if $v$ is a cut node, some template must be applied. 
Since we are matching with template rule, we do not know the
position of the root of expanding nodes.
We enumerate to all possible depths from $1$ to $d-1$ for finding
the root.
For each $\ell \in [1,d)$, we call $\textsc{EncodeP}( v, t, \ell)$
to construct the constraints encoding that $t$ is applied at $v$
and the depth of the root of the expanding nodes is at depth $\ell$.

In $\textsc{EncodeP}$, we can traverse up from $v$ for $\ell$ steps to
find the node to match the root of $t$.
Naturally, it returns $\lfalse$ if there
is no $\ell$th ancestor of $v$. % i.e., there is no match.
The variable $mark$ is the timestamp for node $v$.
% During matching, we will add constraints stating that the matching part of the
% template should match to the nodes added before $mark$
% and expanding part of the template should match with nodes that
% are added at or after $mark$.
The local variable $c$ collects a conjunction of constraints as they are
generated. % and $c$ is initially $\ltrue$.
The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
and traverses the nodes in $t$.
The while loop at~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
the $\ell$th ancestor from $v$,
matches all the ancestors up to the parent of $v$.
As the loop traverses down, it also traverses $t$ along the matching
using variable $v_r$.
In each iteration, $v_r$ is updated to the $j$th child due to lines
\ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
if the ancestors of $v$ also traverse along with the $j$th child.
% (see if condition at line \ref{line:encodep-jth-child-cond}).

% Let us consider the body of the loop.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that
the corresponding node $v_r$ in the template rule is
$MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The loop at line~\ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
We will discuss $\textsc{MatchTree}$ shortly.
% $\textsc{MatchTree}$ is also called with a third parameter
% $mark$ and a Boolean value as the fourth parameter.
% We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get node $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
% After the while loop in $\textsc{EncodeP}$, $v_r$ is the
% candidate node for the root of expanding nodes.
After the while loop in $\textsc{EncodeP}$ at line \ref{line:encodep-vr-expand}, we declare $v_r$
is $Expand$,  match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$, and also call to match the cut pattern at
the subtree of $v$ with the template rule.
